\Large \bf DRAFT: Direct-style process creation on Linux

### 2020-05-17 research/paper/paper.tex
paper

This paper is organized in several sections.
\begin{compactitem}
\item In section \ref{background}, we provide background on fork-style, spawn-style, and direct-style process creation,
and consider their advantages and disadvantages in more depth.
\item In section \ref{overview}, we discuss the design of the API,
including our usage of \texttt{rsyscall}.
\item In section \ref{applications}, we show examples of usage of direct-style process creation in Python on Linux,
including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
\item In section \ref{implementation}, we explore our implementation in depth,
and discuss several difficult process-creation related details of \texttt{rsyscall}'s implementation.
\item In section \ref{evaluation}, we evaluate, somehow (TODO).
\item In section \ref{related_work},
we examine other work which also touches on the topics covered by this paper.
\item In section \ref{future_work},
we list several directions for future work related to direct-style process creation and to \texttt{rsyscall}.
\item In section \ref{conclusions}, we briefly outline our conclusions about direct-style and the future of process creation.
\end{compactitem}
\section{Background}\label{background}
As mentioned in the introduction,
on most operating systems today,
the available interfaces for process creation
are either "fork-style" or "spawn-style".
A few operating systems have a third style, which we refer to as "direct-style".
\subsection{Fork-style}
By far the most well-known fork-style interface is Unix's fork.\cite{forkhist}
With a fork-style interface,
a process, called the parent process, calls \texttt{fork},
and a copy of that process is made,
called the child process,
differing only in a few attributes such as process id.
The same program continues executing in both processes,
with a different return value from fork to allow it to distinguish whether it is running in the parent or the child.
Typically, the program will case on the return value
and, if running in the child,
call various system calls to modify the state of the new process.

@@

For \texttt{execve}, \verb|close_cloexec| is, in some sense, done automatically for us by the kernel;
except we don't get the chance to preserve live file descriptors handles,
without first unsetting \texttt{CLOEXEC} on them.
Again an extension to the kernel would be useful:
A new argument to \texttt{execve} which allows specifying a list of file descriptors to inherit
despite the \texttt{CLOEXEC} flag.
This would allow programs to avoid the pattern
of unsetting \texttt{CLOEXEC}, calling \texttt{execve}, and then immediately re-setting \texttt{CLOEXEC},
just to inherit a file descriptor across a single exec.
This might seem minor,
but unsetting \texttt{CLOEXEC} before \texttt{execve} is not possible
when the process calling \texttt{execve} has a shared file descriptor table,
because some other process might call \texttt{execve} as well,
and erroneously inherit the file descriptor.
\section{Evaluation \textbf{TODO}}\label{evaluation}

We quantitatively answer the following questions:
\begin{compactitem}
\item
  Does direct-style process creation on Linux allow for simpler programs than the alternatives?
  We implement and compare similar functionality using both direct-style and fork-style in \ref{ease}.
\item
  Does a clone-based implementation avoid the well-known performance overheads of fork?
  We evaluate clone against fork on several microbenchmarks in \ref{microbench}.
\item
  Does a high-level Python direct-style process creation interface
  have acceptable performance relative to the Python standard library's spawn-style interface?
  We benchmark direct-style process creation against spawn-style in \ref{subprocess_bench}.
% TODO have an item about use at TS? hm.
% could be qualitative
\end{compactitem}
\subsection{Ease of programming with direct-style}\label{ease}
\begin{table}
\input{ease.tex}
\caption{Line counts with direct-style vs fork-style}
\label{tab:ease}
\end{table}
We implement a variety of simple but powerful techniques with processes
using both direct-style and fork.
Our results are summarized in table \ref{tab:ease}.
Direct-style consistently takes under half the lines the express the same thing.
\lstinputlisting[float=*ht,language=Python,label={basic_fork},caption={Fork-style: Creating a new process, changing CWD, and execing}]
{examples/basic/fork.py}
An example of one of the fork-style implementations is in listing \ref{basic_fork}.
It corresponds to the direct-style example in listing \ref{basic}.

@@

We ensure that errors are reported back to the main program,
even without handling them;
this is a significant source of boilerplate in fork,
because reporting an error back to the main program requires IPC.
Even if there's no desire to handle errors,
proper error reporting is necessary for debugging.
And for serious users, proper error reporting and handling
is essential for a robust application.
% TODO hmm talk about... some stuff.
\subsection{Clone versus fork}\label{microbench}
\begin{figure}[h!]
\centering
 \includegraphics[width=0.5\textwidth]{microbench}
 \caption{fork and clone performance under varying memory usage}
 \label{fig:microbench}
\end{figure}

@@

Furthermore,
our implementation of \texttt{rsyscall} is barely-optimized pure-Python,
all the way down the syscall primitive,
whereas \texttt{subprocess.run} is entirely in C.
If we need to speed up our implementation,
we can perform some targeted optimizations,
reimplement some parts of the Python \texttt{rsyscall} library in native code,
or run our code using a faster Python interpreter such as PyPy.
\section{Related work}\label{related_work}
\subsection{Direct-style process creation}
As we discuss in section \ref{background},
many academic operating systems have direct-style process creation.
We are not aware of any previous instances of direct-style process creation in a Unix-like environment.

@@

Several other aspects of our implementation would also be improved by new features in the Linux kernel,
as discussed in sections \ref{execve} and \ref{cloexec}.
\subsection{Portability to other Unix systems}
Other non-Linux systems
could adopt the techniques of this paper
to provide direct-style process creation.
Currently, our focus is on Linux,
but others may wish to explore porting these techniques to other operating systems.
\subsection{Large scale open source usage}
We have made use of the techniques described in this paper
in proprietary software at \twosigma.
While this gives us personally greater confidence in these techniques,
it would be beneficial to use them in a publicly available, open source system.
Likely the best 
Either porting an existing system to use these techniques,
or using these techniques to create a substantial new system from scratch,
would provide a substantial demonstration of the viability of these techniques.
\section{Conclusions}\label{conclusions}
Direct-style process creation is much less known and much less used than fork-style and spawn-style.
We have implemented direct-style process creation for Linux.
Our implementation is immediately deployable on today's Linux systems.
We have discussed various applications of processes,
and demonstrated the use of Linux direct-style process creation
to implement them.

@@

We hope that a better process creation mechanism
will help encourage more use of the process abstraction,
which, though widespread,
is still not used to its full potential.

>>>>

This paper is organized in several sections.
\begin{compactitem}
\item In section \ref{background}, we provide background on fork-style, spawn-style, and direct-style process creation,
and consider their advantages and disadvantages in more depth.
\item In section \ref{overview}, we discuss the design of the API,
including our usage of \texttt{rsyscall}.
\item In section \ref{applications}, we show examples of usage of direct-style process creation in Python on Linux,
including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
\item In section \ref{implementation}, we explore our implementation in depth,
and discuss several difficult process-creation related details of \texttt{rsyscall}'s implementation.
\item In section \ref{evaluation}, we evaluate using both quantitative and qualitative approaches.
\item In section \ref{related_work},
we examine other work which also touches on the topics covered by this paper.
\item In section \ref{future_work},
we list several directions for future work related to direct-style process creation and to \texttt{rsyscall}.
\item In section \ref{conclusions}, we briefly outline our conclusions about direct-style and the future of process creation.
\end{compactitem}
\section{Background}\label{background}
As mentioned in the introduction,
on most operating systems today,
the available interfaces for process creation
are either "fork-style" or "spawn-style".
A few operating systems have a third style, which we refer to as "direct-style".
\subsection{Fork-style}
By far the most well-known fork-style interface is Unix's fork.\cite{forkhist}
With a fork-style interface,
a process, called the parent process, calls \texttt{fork},
and a copy of that process is made,
called the child process,
differing only in a few attributes such as process id.
The same program continues executing in both processes,
with a different return value from fork to allow it to distinguish whether it is running in the parent or the child.
Typically, the program will case on the return value
and, if running in the child,
call various system calls to modify the state of the new process.

@@

For \texttt{execve}, \verb|close_cloexec| is, in some sense, done automatically for us by the kernel;
except we don't get the chance to preserve live file descriptors handles,
without first unsetting \texttt{CLOEXEC} on them.
Again an extension to the kernel would be useful:
A new argument to \texttt{execve} which allows specifying a list of file descriptors to inherit
despite the \texttt{CLOEXEC} flag.
This would allow programs to avoid the pattern
of unsetting \texttt{CLOEXEC}, calling \texttt{execve}, and then immediately re-setting \texttt{CLOEXEC},
just to inherit a file descriptor across a single exec.
This might seem minor,
but unsetting \texttt{CLOEXEC} before \texttt{execve} is not possible
when the process calling \texttt{execve} has a shared file descriptor table,
because some other process might call \texttt{execve} as well,
and erroneously inherit the file descriptor.
\section{Evaluation}\label{evaluation}
We answer the following questions:
\begin{compactitem}
\item
  Does direct-style process creation on Linux allow for simpler programs than the alternatives?
  We implement and compare similar functionality using both direct-style and fork-style in \ref{ease}.
\item
  Does a clone-based implementation avoid the well-known performance overheads of fork?
  We evaluate clone against fork on several microbenchmarks in \ref{microbench}.
\item
  Does a high-level Python direct-style process creation interface
  have acceptable performance relative to the Python standard library's spawn-style interface?
  We benchmark direct-style process creation against spawn-style in \ref{subprocess_bench}.
\item
  Does direct-style process creation work in the ``real world''?
  We discuss our experience with using direct-style process creation at \twosigma in \ref{realworld}.
\end{compactitem}
\subsection{Ease of programming with direct-style}\label{ease}
\begin{table}
\input{ease.tex}
\caption{Line counts with direct-style vs fork-style}
\label{tab:ease}
\end{table}
We implement a variety of simple but powerful techniques with processes
using both direct-style and fork.
Our results are summarized in table \ref{tab:ease}.
Direct-style consistently takes under half the lines the express the same thing.
\lstinputlisting[float=*ht,language=Python,label={basic_fork},caption={Fork-style: Creating a new process, changing CWD, and execing}]
{examples/basic/fork.py}
An example of one of the fork-style implementations is in listing \ref{basic_fork}.
It corresponds to the direct-style example in listing \ref{basic}.

@@

We ensure that errors are reported back to the main program,
even without handling them;
this is a significant source of boilerplate in fork,
because reporting an error back to the main program requires IPC.
Even if there's no desire to handle errors,
proper error reporting is necessary for debugging.
And for serious users, proper error reporting and handling
is essential for a robust application.
\subsection{Clone versus fork}\label{microbench}
\begin{figure}[h!]
\centering
 \includegraphics[width=0.5\textwidth]{microbench}
 \caption{fork and clone performance under varying memory usage}
 \label{fig:microbench}
\end{figure}

@@

Furthermore,
our implementation of \texttt{rsyscall} is barely-optimized pure-Python,
all the way down the syscall primitive,
whereas \texttt{subprocess.run} is entirely in C.
If we need to speed up our implementation,
we can perform some targeted optimizations,
reimplement some parts of the Python \texttt{rsyscall} library in native code,
or run our code using a faster Python interpreter such as PyPy.
\subsection{Usage in the real world}\label{realworld}
At \twosigma we have used \texttt{rsyscall} and direct-style process creation
as a component of a distributed system deployment library.
This library is in use in production,
and is extensively used as part of testing and development.

With direct-style process creation,
we can robustly and easily write custom code to start up each program,
passing down file descriptors and initializing process attributes according to each program's needs,
all of which can be done directly from a single parent process.
This has allowed us to make more use of file descriptor abstraction;
for example, connections over our shared memory transport are bootstrapped over passed-down file descriptors,
and our management interfaces listen on passed-in sockets,
allowing them to be easily swapped between network-accessible TCP sockets and host-local Unix domain sockets.
We can also flexibly use namespaces where appropriate;
for example, for legacy applications which directly start subprocesses,
we prevent process leaking with pid namespaces.

All this has allowed us to run our systems in a much more dynamic way than before;
we can freely start up arbitrary subsets of our systems on-demand,
spread across one or more hosts,
for development, testing, or production,
regardless of the configuration of that host,
without worrying about interference between instances,
and without a dependency on any external privileged services,
such as container runtimes.
This has become a key part of our new development workflow.
\section{Related work}\label{related_work}
\subsection{Direct-style process creation}
As we discuss in section \ref{background},
many academic operating systems have direct-style process creation.
We are not aware of any previous instances of direct-style process creation in a Unix-like environment.

@@

Several other aspects of our implementation would also be improved by new features in the Linux kernel,
as discussed in sections \ref{execve} and \ref{cloexec}.
\subsection{Portability to other Unix systems}
Other non-Linux systems
could adopt the techniques of this paper
to provide direct-style process creation.
Currently, our focus is on Linux,
but others may wish to explore porting these techniques to other operating systems.
\subsection{Large scale open source usage}
We would like to open source libraries built on top of direct-style process creation.
From our experience using such libraries at \twosigma,
we believe this would greatly ease the development of complex systems involving processes,
and would provide a substantial demonstration of the power of direct-style process creation.
\section{Conclusions}\label{conclusions}
Direct-style process creation is much less known and much less used than fork-style and spawn-style.
We have implemented direct-style process creation for Linux.
Our implementation is immediately deployable on today's Linux systems.
We have discussed various applications of processes,
and demonstrated the use of Linux direct-style process creation
to implement them.

@@

We hope that a better process creation mechanism
will help encourage more creative use of processes,
which, though widespread,
are still not used to their full potential.
